---
title: "Predictive Analytics: Problem Sets 6 & 7"
author: "Meghna Krishan - 749"
output: html_document
---
```{r}
library(ISLR)
library(leaps)
```

# Problem Set 6: Model Selection (Subset Selection Methods)
```{r}
# 1. Clean data - remove rows with missing values
data(Hitters)
Hitters <- na.omit(Hitters)
```


```{r}
# 2. Train-test split (80-20)
set.seed(42)
n <- nrow(Hitters)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- Hitters[train_idx, ]
test  <- Hitters[-train_idx, ]
```

```{r}
# Helper function: compute MSE from regsubsets model
predict_regsubsets <- function(object, newdata, id) {
  form  <- as.formula(object$call[[2]])
  mat   <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  mat[, names(coefi)] %*% coefi
}

get_mse <- function(object, traindata, testdata, nvars) {
  train_mat <- model.matrix(Salary ~ ., traindata)
  test_mat  <- model.matrix(Salary ~ ., testdata)
  
  sapply(nvars, function(i) {
    coefi      <- coef(object, id = i)
    train_pred <- train_mat[, names(coefi)] %*% coefi
    test_pred  <- test_mat[, names(coefi)]  %*% coefi
    c(train_mse = mean((traindata$Salary - train_pred)^2),
      test_mse  = mean((testdata$Salary  - test_pred)^2))
  })
}

# Function to select best model per criterion and report MSEs
report_best <- function(reg_obj, train_df, test_df, method_name) {
  s <- summary(reg_obj)
  
  best_adjr2 <- which.max(s$adjr2)
  best_bic   <- which.min(s$bic)
  best_cp    <- which.min(s$cp)
  # AIC not directly from leaps; approximate via BIC-like formula or use Cp (proportional)
  # AIC ~ n*log(RSS/n) + 2p; we compute manually
  n_tr <- nrow(train_df)
  aic_vals <- n_tr * log(s$rss / n_tr) + 2 * (1:19 + 1)
  best_aic  <- which.min(aic_vals)
  
  criteria <- c(AdjR2 = best_adjr2, BIC = best_bic, AIC = best_aic, Cp = best_cp)
  cat("\n===", method_name, "===\n")
  cat("Best model size by criterion:\n"); print(criteria)
  
  results <- data.frame(Criterion = names(criteria), Best_Size = as.integer(criteria),
                        Train_MSE = NA, Test_MSE = NA)
  
  train_mat <- model.matrix(Salary ~ ., train_df)
  test_mat  <- model.matrix(Salary ~ ., test_df)
  
  for (i in seq_len(nrow(results))) {
    k     <- results$Best_Size[i]
    coefs <- coef(reg_obj, id = k)
    tr_p  <- train_mat[, names(coefs)] %*% coefs
    te_p  <- test_mat[, names(coefs)]  %*% coefs
    results$Train_MSE[i] <- mean((train_df$Salary - tr_p)^2)
    results$Test_MSE[i]  <- mean((test_df$Salary  - te_p)^2)
  }
  
  print(results)
  invisible(results)
}
```

## 3-4. Best Subset Selection

```{r}
reg_best <- regsubsets(Salary ~ ., data = train, nvmax = 19, method = "exhaustive")
res_best <- report_best(reg_best, train, test, "Best Subset Selection")
```

## 5. Forward Stepwise Selection

```{r}
reg_fwd <- regsubsets(Salary ~ ., data = train, nvmax = 19, method = "forward")
res_fwd <- report_best(reg_fwd, train, test, "Forward Stepwise Selection")
```

## 6. Backward Stepwise Selection

```{r}
reg_bwd <- regsubsets(Salary ~ ., data = train, nvmax = 19, method = "backward")
res_bwd <- report_best(reg_bwd, train, test, "Backward Stepwise Selection")
```

## 7. Comparison of Three Methods

```{r}
cat("\n=== Summary Comparison ===\n")
cat("\nBest Subset:\n");  print(res_best)
cat("\nForward Stepwise:\n"); print(res_fwd)
cat("\nBackward Stepwise:\n"); print(res_bwd)
```

> **Interpretation:** All three methods tend to select similar models for most criteria (especially BIC and Cp which favor parsimony). Forward and Backward stepwise are computationally cheaper than Best Subset but may differ slightly in selected variables. Test MSE is the most reliable guide to generalization performance.

---

# Problem Set 7: Shrinkage Methods

```{r}
if (!require(glmnet)) install.packages("glmnet")
library(glmnet)

# Fresh clean data
data(Hitters)
Hitters <- na.omit(Hitters)

# 1. Split: 100 test, rest training
set.seed(42)
test_idx  <- sample(1:nrow(Hitters), 100)
train7    <- Hitters[-test_idx, ]
test7     <- Hitters[test_idx, ]

X_train <- model.matrix(Salary ~ ., train7)[, -1]
y_train <- train7$Salary
X_test  <- model.matrix(Salary ~ ., test7)[, -1]
y_test  <- test7$Salary

```

## 2. Least Squares

```{r}
ols_fit   <- lm(Salary ~ ., data = train7)
ols_train <- predict(ols_fit, train7)
ols_test  <- predict(ols_fit, test7)

ols_train_mse <- mean((y_train - ols_train)^2)
ols_test_mse  <- mean((y_test  - ols_test)^2)
```

## 3. Ridge Regression

```{r}
# Grid of lambda values
lambda_grid <- 10^seq(10, -2, length = 100)

ridge_fit <- glmnet(X_train, y_train, alpha = 0, lambda = lambda_grid)

# (i) Dimension of coefficient matrix
cat("Coefficient matrix dimensions:", dim(coef(ridge_fit)), "\n")

# (ii) L2 norm at 50th and 60th lambda
l2_50 <- sqrt(sum(coef(ridge_fit)[-1, 50]^2))
l2_60 <- sqrt(sum(coef(ridge_fit)[-1, 60]^2))
cat("L2 norm at lambda[50] =", round(lambda_grid[50], 4), ":", round(l2_50, 4), "\n")
cat("L2 norm at lambda[60] =", round(lambda_grid[60], 4), ":", round(l2_60, 4), "\n")
cat("Larger lambda => smaller L2 norm (more shrinkage)\n")

# (iv) Test MSE for 50th and 60th lambda
pred_50 <- predict(ridge_fit, s = lambda_grid[50], newx = X_test)
pred_60 <- predict(ridge_fit, s = lambda_grid[60], newx = X_test)
cat("Test MSE at lambda[50]:", round(mean((y_test - pred_50)^2), 2), "\n")
cat("Test MSE at lambda[60]:", round(mean((y_test - pred_60)^2), 2), "\n")
```
