---
title: "Problem Set 3 Q3"
author: "Meghna"
date: "2026-02-26"
output: html_document
---

**Question 3: *Problem to demonstrate the role of qualitative (ordinal) predictors in addition to quantitative predictors in multiple linear regression.***

Consider “diamonds” data set in R. It is in the ggplot2 package. Make a list of all the ordinal categorical variables. Identify the response.

```{r}
library(ggplot2)
data("diamonds")
head(diamonds)
```

ANSWER: The ordinal categorical variables in the diamonds dataset are "cut", "color", and "clarity". The response variable is "price".

(a) Run a linear regression of the response on the quality of cut. Write the fitted regression model.

```{r}
model_cut <- lm(price ~ cut, data = diamonds)
```

(b) Test whether the expected price of diamond with premium cut is significantly different from that of the ideal cut.

```{r}
summary(model_cut)
```

(c) What is the expected price of a diamond of ideal cut?

```{r}
ideal_price <- predict(model_cut, newdata = data.frame(cut = "Ideal"))
ideal_price
```

(d) Modify the regression model in (a) by incorporating the predictor “table”. Write the fitted regression model.

```{r}
model_cut_table <- lm(price ~ cut + table, data = diamonds)
model_cut_table
```

(e) Test for the significance of “table” in predicting the price of diamond.

```{r}
summary(model_cut_table)
```

(f) Find the average estimated price of a diamond with an average table value and which is of fair cut.

```{r}
average_table <- mean(diamonds$table)
average_table
pred= predict(model_cut_table,newdata = data.frame(cut="Fair",table=average_table));pred
```
